---
title: "Leakage-aware features and subject-safe splits"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Leakage-aware features and subject-safe splits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

PhysioML turns [`PhysioExperiment`](https://x-biosignal.github.io/PhysioExperiment/)
objects into model-ready feature matrices while keeping the two things that are
easy to get wrong in physiological machine learning explicit and auditable:

* **case-to-subject identity**, so a split generalises to *unseen people* rather
  than to unseen trials of people the model already saw, and
* **transform provenance**, so a convolution transform fitted on training cases
  is reused unchanged on held-out cases.

This vignette walks the fully offline parts of that workflow. The deep-learning
(torch/luz), ROCKET (Python/aeon), ONNX, and foundation-model paths are
backend-gated and are shown as (unevaluated) code at the end.

## A small experiment

We start from a `PhysioExperiment` whose assay is a `time x channel x case`
array. Each case is one labelled trial; several cases may come from the same
subject.

```{r setup}
library(PhysioML)
library(PhysioExperiment)
library(S4Vectors)

set.seed(42)
n_case <- 8
arr <- array(stats::rnorm(60 * 2 * n_case), dim = c(60, 2, n_case))
pe <- PhysioExperiment(
  assays = list(raw = arr),
  colData = DataFrame(label = c("C3", "C4")),
  samplingRate = 100
)
pe
```

## Attach subject identity, then split on subjects

A cycle or trial is a *case*, not an independent subject. `withCaseData()`
records an explicit case-to-subject mapping instead of guessing subjects from
case names. Cases in this assay without names are addressed as `case_1`,
`case_2`, ...

```{r identity}
case_data <- data.frame(
  case_id    = paste0("case_", seq_len(n_case)),
  subject_id = paste0("subj_", rep(1:4, each = 2))
)
pe <- withCaseData(pe, case_data)
```

`validateSubjectSplit()` then refuses a split that shares a case *or a subject*
between training and validation -- the silent leak that inflates reported
accuracy.

```{r split}
train <- case_data[case_data$subject_id %in% c("subj_1", "subj_2"), ]
valid <- case_data[case_data$subject_id %in% c("subj_3", "subj_4"), ]
validateSubjectSplit(train, valid)
```

## Feature extraction

`peReducedFeatures()` produces a case-by-feature design matrix. The canonical
catch22 characteristics are computed through the official `Rcatch22` backend, so
that step runs only when the optional package is installed:

```{r features}
if (requireNamespace("Rcatch22", quietly = TRUE)) {
  features <- peReducedFeatures(pe, method = "catch22")
  dim(features)
} else {
  message("Install Rcatch22 to compute catch22 features.")
}
```

The convolution transforms (`rocket()`, `minirocket()`) follow the same
contract but use a caller-managed Python/aeon backend. When transforming
held-out cases, pass the model fitted only on training cases so no information
leaks from validation into the transform.

## Aligning two domains with CORAL

`domainAdapt()` matches the first- and second-order statistics of a labelled
source domain to an unlabelled target domain. It is pure R and runs offline. Fit
it with training-source observations and a designated unlabelled adaptation set
only -- never with held-out target data.

```{r coral}
feat <- c("f1", "f2", "f3")
source <- matrix(stats::rnorm(30), ncol = 3, dimnames = list(NULL, feat))
target <- matrix(stats::rnorm(30, mean = 1), ncol = 3,
                 dimnames = list(NULL, feat))
fit <- domainAdapt(source, target)
fit$diagnostics$source_mean_residual
```

The residual is ~0 because aligned source means now match the target means.

## Backend-gated workflows

The remaining workflows need optional backends and are shown here for
orientation only (not evaluated in this vignette):

```{r dl, eval = FALSE}
# Deep learning (needs torch + luz)
train_ds <- peDataset(
  pe, targets = rep(c("a", "b"), 4),
  task = "classification", class_levels = c("a", "b")
)
model <- trainModel(train_ds, model = "cnn1d", epochs = 10L)
predictModel(model, train_ds, type = "class")

# Governed ONNX round-trip (needs the torch -> ONNX bridge + ONNX Runtime)
path <- tempfile(fileext = ".onnx")
exportONNX(model, path)
onnx_model <- importONNX(path)

# Governed foundation-model embeddings (needs a Python env + pinned revision)
foundationEmbed(
  pe, model = "moment",
  revision = "411e288267f82cce86296dbe4d6c8bc533cc162f",
  manifest_sha256 = "<expected manifest hash>",
  allow_download = TRUE
)
```

None of these mechanics establish clinical validity; they establish
*reproducibility* and *leakage discipline*. See `?peReducedFeatures`,
`?validateSubjectSplit`, and `?trainModel` for the per-function contracts.
